A Simple Type Theory for Category Theory

### 2022-02-06 paper/lipics-v2021-sample-article.tex
synthetic progress

\newcommand{\odotexists}[1]{\mathop{\overset{\exists #1}\odot}}
\newcommand{\tensorexistsXwithYandZ}[3]{#2 \odotexists{#1} #3}
\newcommand{\tensorintroofXandY}[2]{\textrm{pair}_\odot(#1,#2)}
\newcommand{\tensorelimXtoYwithkontZ}[3]{\textrm{ind}_{\to}(#1,#2,#3)}
\newcommand{\tensorelimfromWXtoYwithkontZ}[4]{\textrm{ind}_{\odot}(#1,#2,#3,#4)}
\newcommand{\tensorelimVatWtoXtoYwithkontZ}[5]{\textrm{ind}_{\to}(#4,#2,#1,#3,#5)}

@@

\newcommand{\homlallXYtoZ}[3]{#3 \tlforall #1 #2}

@@

\newcommand{\homunary}[4]{\pendallXdotY {#1} {\homrallXYtoZ {#3} {#2} {#4}}}

@@

\begin{figure}
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Type} & A,B,C & ::= &
      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      &&& \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N & ::= & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M\\
      &&&\lceil \cat C \rceil \pipe \vlambda {\alpha:\cat C} a \pipe \hlambda {\alpha:\cat C}{\beta: \cat D} R \ldots \\
      \textrm{Categories} &\cat C, \cat D, \cat E & ::= & M \pipe \cat C \times \cat D \pipe \cat 1\\
      \textrm{Objects} & a, b, c & ::= & \alpha \pipe M a \pipe (a,b) \pipe () \ldots\\
      \textrm{Sets} & P,Q,R & ::= & \punitinXfromYtoZ {\cat C} a b \pipe \tensorexistsXwithYandZ {\beta} P Q \pipe \homrallXYtoZ \beta P Q \pipe \homlallXYtoZ \alpha R S \\
      &&& M(a,b) \pipe 1 \pipe P \times Q\\
      \textrm{Transformations} & s,t,u & ::= & \ldots \\
      \\
      \textrm{Type Context} & \Gamma, \Delta & ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Category Context} & \Xi & ::= & \alpha:\cat C \pipe \alpha:\cat C;\beta: \cat D\\
      \textrm{Set Context} & \Phi,\Psi & ::= & \alpha : \cat C \pipe \Phi , x : P , \Psi\\
      \textrm{Term Substitution} & \gamma, \delta & ::= & \cdot \pipe \gamma, M/X\\
      \textrm{Object Substitution} & \xi, \zeta & ::= & a/\alpha \pipe a/\alpha;b/\beta\\
      \textrm{Transformation Substitution} & \phi, \psi & ::= & a/\alpha \pipe \phi,s/x, \psi      
    \end{array}
  \end{mathpar}
  \caption{Simple Category Theory Syntactic Forms}
  \label{fig:syntax}
\end{figure}

@@

First, the multiplicative unit is the set $\punitinXfromYtoZ {\cat C}
a {a'}$, that is the hom set of the category $\cat C$. The rules for
this type act like .

we can define the
\emph{unit} , which in set-theoretic settings we
might call the \emph{identity/equality judgment} $A \equiv_{\cat C}
A'$.
%
First the formation rule says given two objects $A,A'$ in the same
category $\cat C$, we can form the hom-set from $A$ to $A'$ $A
\to_{\cat C} A'$.
%
The introduction rule says that given any object $A$, we can define
the identity arrow from $A$ to itself $\id(A) : A \to_{\cat C} A$.
%
Without knowing more about the category in question, these are the
only arrows we know exist.
%

\begin{figure}
  \begin{mathpar}
    \inferrule*[right=Unit-Formation]
    {\Gamma \vdash C : \Sort
     \Gamma \pipe d^-\Xi \vdash A : \cat C\and \Gamma \pipe d^+ \Xi \vdash A' : \cat C}
    {\Gamma \pipe \Xi \vdash \punitinXfromYtoZ {C} A {A'}}

@@

    \inferrule*[right=Unit-Introduction]
    {\Gamma\pipe\cdot\vdash C : \Sort}
    {\Gamma \pipe\cdot \vdash \punitrefl C : \pendallXdotY {\alpha:C} \punitinXfromYtoZ C \alpha {\alpha}}

    \inferrule*[right=Unit Elim]{
      \Gamma\pipe\cdot \vdash Q : \harr C C\and
      \Gamma\pipe\cdot\vdash N : \pendallXdotY {\alpha:C} {\harrapp Q \alpha \alpha}}
    {\Gamma\pipe\cdot \vdash \punitelimtoYwithkontZ Q N :
      \pendallXdotY {\alpha} {\homrallXYtoZ {\alpha'} {\punitinXfromYtoZ C {\alpha}{\alpha'}} {\harrapp Q {\alpha}{\alpha'}}}
      }

@@

    \homrappXtoYatZ {\pendappXtoY {\punitelimtoYwithkontZ Q N} A} {\pendappXtoY {\punitrefl C} A} A = \pendappXtoY N A

    \inferrule*[right=Unit$\eta$]
    {\alpha_1 : \cat C, f : \alpha_1 \to_{\cat C} \alpha_2, \alpha_2:\cat C \vdash M : \harrapp Q {\alpha_1} {\alpha_2}}
    {\alpha_1 : \cat C, f : \alpha_1 \to_{\cat C} \alpha_2, \alpha_2:\cat C \vdash M =
      \homrappXtoYatZ {\pendappXtoY {\punitelimtoYwithkontZ Q {\pendlambdaXdotY \alpha M[\alpha/\alpha_1,\pendappXtoY {\punitrefl C} \alpha/f,\alpha/\alpha_2]}} {\alpha_1}} f {\alpha_2}
    : \harrapp Q {\alpha_1}{\alpha_2}}
  \end{mathpar}  
  \caption{Unit Primitive Rules}
\end{figure}

Next, we can internalize the $,\alpha,$ as a positive connective as a
kind of ``existentially quantified tensor product'', abstracting the
composition of profunctors.

The ordinary formula for this is that for profunctors $R : C^o \times
D \to \Set$ and $S : D^o \times E \to \Set$, the composite is given by
a coend:

\[ \int^{d \in D} R(-,d) \otimes S(d,=) \]

We will instead use slightly more logical notation, writing the coend as
an existential quantifier, justified by the analogy between the two
and the structure of the introduction/elimination forms in our syntax
in particular.
\begin{figure}
\begin{mathpar}
  \inferrule*[right=Tensor formation]
  {\Gamma \pipe d^-\Xi; \beta:D\vdash P \isaSet\and
    \Gamma \pipe \beta:D;d^+\Xi \vdash Q}
  {\Gamma \pipe \Xi \vdash \tensorexistsXwithYandZ {\beta:D} P Q}

@@

  \inferrule*[right=Tensor introduction]
  {\Gamma \vdash P : \harr C D \and
    \Gamma \vdash Q : \harr D E}
  {\Gamma\pipe\cdot\vdash \tensorintroofXandY P Q : \pendallXdotY {\alpha} \homrallXYtoZ \beta {\harrapp P \alpha \beta} {\homrallXYtoZ \gamma {\harrapp Q \beta \gamma} {(\tensorexistsXwithYandZ \beta {\harrapp P \alpha \beta}{\harrapp Q \beta \gamma})}}}

  %% \inferrule*[right=Tensor introduction]
  %% {\Phi = \Psi_M \jnctx \Psi_N \and
  %%  \Gamma \pipe d^+\Psi_M \vdash B : D\and
  %%  \Gamma \pipe \Psi_M \vdash M : P[;B/\beta] \and
  %%  \Gamma \pipe \Psi_N \vdash N : Q[B/\beta;]}
  %% {\Gamma\pipe\Phi\vdash \tensorintroatXwithYandZ B M N : \tensorexistsXwithYandZ {\beta : D} P Q}

@@

  \inferrule*[right=Tensor-elim]
  {\Gamma \pipe \cdot \vdash P : \harr C D\and\Gamma \pipe \cdot \vdash Q : \harr D E\and
   \Gamma \pipe \cdot \vdash R : \harr C E\and
   \Gamma \pipe \cdot \vdash N : \pendallXdotY \alpha {\homrallXYtoZ \beta {\harrapp P \alpha \beta} {\homrallXYtoZ \gamma {\harrapp Q \beta \gamma}}}   
  }
  {\Gamma\pipe\cdot \vdash \tensorelimfromWXtoYwithkontZ P Q R N : \pendallXdotY \alpha {\homrallXYtoZ \gamma {\tensorexistsXwithYandZ \beta {\harrapp P \alpha \beta}{\harrapp Q \beta \gamma}} {\harrapp R \alpha \gamma}}}

@@

  \inferrule*[right=Tensor $\beta$]
  {}
  {\homrappXtoYatZ {\pendappXtoY{\tensorelimfromWXtoYwithkontZ P Q R N} A} {(\homrappXtoYatZ{\homrappXtoYatZ{\pendappXtoY {\tensorintroofXandY P Q} A} f B} g G)} G =
    \homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY N A} f B} g G}

@@

  \inferrule*[right=Tensor$\eta$]
  {\alpha : \cat C, x : \tensorexistsXwithYandZ \beta P Q, \gamma : \cat E \vdash M : R}
  {\alpha : \cat C, x : \exists \beta. Q \odot R, \gamma : \cat E \vdash M =
    \homrappXtoYatZ {\pendappXtoY{\tensorelimfromWXtoYwithkontZ P Q R
        {\pendlambdaXdotY \alpha {\homrlambdaXatYdotZ p \beta {\homrlambdaXatYdotZ q \gamma {M[\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {\tensorintroofXandY P Q} \alpha} p \beta} q \gamma/x]}}}}
      } \alpha} x \gamma : S}
\end{mathpar}
\caption{Tensor Primitive Rules}
\label{fig:tensor}
\end{figure}
First, the syntax for tensor is based on the coend formula that gives
the definition in $Cat$, but using logical syntax $\exists$ rather
than integrals because I don't like integral syntax and the syntax
matches existential quantification in type theory.
%
Note that $\exists \beta. ~Q \odot R$ is the application of a single
connective to $Q$ and $R$, and not separable into a quantifier
$\exists$ connective and a product connective $\odot$.
%
Introduction, elimination and $\beta\eta$ are similar to a combination
of existential quantification and the tensor product in linear logic.
%
The introduction rule is justified by the fact that semantically what
we have is a transformation
\[ \textrm{pair} : (\alpha : \cat C,Q,\beta:\cat D,R,\gamma:\cat E) \Rightarrow \exists \beta. R \odot Q \]
And the introduction rule is just building in a substitution to this
transformation.
%

@@

\subsection{Equalizers}

To encode the concept of internal hom of profunctors, kan extensions
and (co-)limits, we need to define the \emph{equalizer} of horizontal
arrows. Specifically, the universal property that we need is an
equalizer in the category whose objects are horizontal arrows (for
fixed endpoints) and whose arrows are the 2-cells.

\begin{figure}
  \begin{mathpar}
    \inferrule*[right=Equalizer-formation]
    {\Gamma \pipe \cdot \vdash P : \harr\, C\,D\and
     \Gamma \pipe \cdot \vdash Q : \harr\, C \, D\\\\
     \Gamma \pipe \cdot \vdash \phi : \homunary \alpha {\harrapp P \alpha \beta} \beta {\harrapp Q \alpha \beta}\and
     \Gamma \pipe \cdot \vdash \psi : \homunary \alpha {\harrapp P \alpha \beta} \beta {\harrapp Q \alpha \beta}\and}
    {\Gamma \pipe \cdot \vdash \textrm{Equalizer}(C,D,P,\phi,\psi) : \harr\, C\, D}\and

    \inferrule
    {\Gamma \pipe \Phi \vdash M : \harrapp{\textrm{Equalizer}(C,D,P,\phi,\psi)} A B}
    {\Gamma \pipe \Phi \vdash \textrm{forget}(M) : \harrapp P A B}

    \inferrule
    {\Gamma \pipe \Phi \vdash M : \harrapp P A B \and \appunary \phi A M B = \appunary \psi A M B}
    {\Gamma \pipe \Phi \vdash \textrm{equalizing}_{\phi,\psi}(M) : \harrapp{\textrm{Equalizer}(C,D,P,\phi,\psi)} A B}

    TODO: obvious equations
  \end{mathpar}
  \caption{Equalizers (Logical Framework style)}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule*[right=Equalizer-formation]
    {\Gamma \pipe \Xi \vdash P \isaSet\and\Gamma \pipe \Xi \vdash Q \isaSet\and
      \Gamma\pipe d^-\Xi, p : P, d^+ \Xi \vdash \phi : Q\and
      \Gamma\pipe d^-\Xi, p : P, d^+ \Xi \vdash \psi : Q }
    {\Gamma \pipe \Xi \vdash \equalizeVofWbyXeqYatZ p P \phi \psi Q : \harr\, C\, D}\and

    \inferrule*[right=Equalizer-formation]
    {\Gamma \pipe \alpha:C;\beta:D \vdash P \isaSet\and\Gamma \pipe \alpha:C;\beta:D \vdash Q \isaSet\and
      \Gamma \pipe d^-\Xi \vdash A : C\and
      \Gamma \pipe d^+\Xi \vdash B : D\\\\
      \Gamma\pipe \cdot \vdash \phi : \homunary \alpha P \beta Q\and
      \Gamma\pipe \cdot \vdash \psi : \homunary \alpha P \beta Q
    }
    {\Gamma \pipe \Xi \vdash \equalizeVofWbyXeqYatZ p {P[A/\alpha;B/\beta]} {\appunary \phi A p B} {\appunary \psi A p B} {Q[A/\alpha;B/\beta]} : \harr\, C\, D}\and

    \inferrule
    {\Gamma \pipe \Phi \vdash M : \equalizeVofWbyXeqYatZ p P \phi \psi Q}
    {\Gamma \pipe \Phi \vdash \pi_{eq}\,M : P}

    \inferrule
    {\Gamma \pipe \Phi \vdash M : \harrapp P A B \and \appunary \phi A M B = \appunary \psi A M B}
    {\Gamma \pipe \Phi \vdash \textrm{equalizing}_{\phi,\psi}(M) : \harrapp{\textrm{Equalizer}(C,D,P,\phi,\psi)} A B}

    TODO: equations
  \end{mathpar}
  \caption{Equalizers (Natural Deduction Style)}
\end{figure}

\subsection{Products}

We add in parallel two different notions of product: products of sorts
and products of spans, which have the universal property in their
respective categories.

\begin{figure}
  \begin{mathpar}
    \inferrule{}{\Gamma \vdash 1 : \Sort}\and
    \inferrule*{}{\Gamma\pipe\alpha:C \vdash () : 1}\and
    \inferrule*{\Gamma\pipe\alpha:C \vdash A : 1}{A = ()}
    \inferrule{}{\Gamma\pipe \Xi \vdash 1 \isaSet}\and
    \inferrule{}{\Gamma\pipe \Phi \vdash () :  1}\and
    \inferrule{\Gamma\pipe \Phi \vdash M :  1}{M = ()}\\\\

    \inferrule{\Gamma \vdash C : \Sort \and \Gamma \vdash D : \Sort}
              {\Gamma \vdash C \times D : \Sort}\and

    \inferrule{\forall i \in \{1,2\}.~ \Gamma \pipe \alpha:C \vdash A_i : D_i}{\Gamma \pipe \alpha:C \vdash (A_1,A_2) : D_1 \times D_2}

    \inferrule{\Gamma \pipe \alpha:C \vdash A : D_1 \times D_2}{\Gamma \pipe \alpha:C \vdash \pi_i A : D_i}

    \pi_i(A_1,A_2) = A_i\\
    \inferrule{\Gamma\pipe \alpha:C \vdash A :  D_1\times D_2}{A = (\pi_1 A,\pi_2 A)}\\\\

    \inferrule{\Gamma\pipe\Xi \vdash R \isaSet \and \Gamma\pipe\Xi \vdash S\isaSet}
              {\Gamma\pipe\Xi \vdash R \times S\isaSet}\and

    \inferrule{\forall i \in \{1,2\}.~ \Gamma \pipe\Phi \vdash M_i : R_i}{\Gamma \pipe\Phi \vdash (M_1,M_2) : R_1 \times R_2}

    \inferrule{\Gamma \pipe\Phi\vdash M : R_1 \times R_2}{\Gamma \pipe\Phi \vdash \pi_i M : R_i}

>>>>

\newcommand{\odotexists}[1]{\mathop{\overset{\exists #1}\odot}}
\newcommand{\tensorexistsXwithYandZ}[3]{#2 \odotexists{#1} #3}
\newcommand{\tensorintroofXandY}[2]{\textrm{pair}_\odot(#1,#2)}
\newcommand{\tensorintroatXwithYandZ}[3]{(#2,#1,#3)}
\newcommand{\tensorelimXtoYwithkontZ}[3]{\textrm{ind}_{\to}(#1,#2,#3)}
\newcommand{\tensorelimfromWXtoYwithkontZ}[4]{\textrm{ind}_{\odot}(#1,#2,#3,#4)}
\newcommand{\tensorelimVatWtoXtoYwithkontZ}[5]{\textrm{ind}_{\to}(#4,#2,#1,#3,#5)}

@@

\newcommand{\homlallXYtoZ}[3]{#3 \tlforall {#1} #2}

@@

\newcommand{\homunary}[4]{\pendallXdotY {#1} {\homrallXYtoZ {#3} {#2} {#4}}}
\newcommand{\hombinary}[6]{\pendallXdotY {#1} {\homrallXYtoZ {#3} {#2} \homrallXYtoZ {#5} {#4} {#6}}}

@@

\begin{figure}
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Type} & A,B,C & ::= &
      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      &&& \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N & ::= & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M\\
      &&&\lceil \cat C \rceil \pipe \vlambda {\alpha:\cat C} a \pipe \hlambda {\alpha:\cat C}{\beta: \cat D} R \ldots \\
      \textrm{Categories} &\cat C, \cat D, \cat E & ::= & M \pipe \cat C \times \cat D \pipe \cat 1\\
      \textrm{Objects} & a, b, c & ::= & \alpha \pipe M a \pipe (a,b) \pipe () \ldots\\
      \textrm{Sets} & P,Q,R & ::= & \punitinXfromYtoZ {\cat C} a b \pipe \tensorexistsXwithYandZ {\beta} P Q \pipe \homrallXYtoZ \beta P Q \pipe \homlallXYtoZ \alpha R S \\
      &&& M(a,b) \pipe 1 \pipe P \times Q\\
      \textrm{Transformations} & s,t,u & ::= & \ldots \\
      \\
      \textrm{Type Context} & \Gamma, \Delta & ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Category Context} & \Xi, Z & ::= & \alpha:\cat C \pipe \alpha:\cat C;\beta: \cat D\\
      \textrm{Set Context} & \Phi,\Psi & ::= & \alpha : \cat C \pipe \Phi , x : P , \Psi\\
      \textrm{Term Substitution} & \gamma, \delta & ::= & \cdot \pipe \gamma, M/X\\
      \textrm{Object Substitution} & \xi, \zeta & ::= & a/\alpha \pipe a/\alpha;b/\beta\\
      \textrm{Transformation Substitution} & \phi, \psi & ::= & a/\alpha \pipe \phi,s/x, \psi      
    \end{array}
  \end{mathpar}
  \caption{Simple Category Theory Syntactic Forms}
  \label{fig:syntax}
\end{figure}

@@

First, the multiplicative unit is the set $\punitinXfromYtoZ {\cat C}
{a_1} {a_2}$, semantically, the hom set of the category $\cat C$. The
rules for this type act like an identity type in a directed type
theory. The introduction rule defines an identity morphism of an
object in any category. The elimination rule is then the appropriate
analogue of the $J$ rule of the identity type. Given any profunctor
$Q$ from $\cat C$ to itself, we can construct an element $\harrapp Q
{\alpha_1}{\alpha_2}$ for any objects in $\cat C$ if we specify a
natural element $\harrapp Q \alpha \alpha$ for any $\alpha$ in $\cat
C$. The reason this is sufficient, semantically, is that all
profunctors must be functorial in their two indices. In fact, as we
will see in the section on synthetic category theory, this induction
principle is how we can show internally that profunctors are
functorial in their indices. We choose to express this elimination
form using the right hom type as its syntax more closely matches the
usual syntax of function types, but this choice is arbitrary, and we
would get an equivalent calculus had we used the left hom, or added
both.

Next, we continue with the tensor product $\tensorexistsXwithYandZ
\beta P Q$. Adjoint to the right/left hom sets, this provides a form
of \emph{existentially quantified pair}. The formation rule states the
set is well-formed when $\beta$ is the covariant variable of $P$ and
the contravariant variable of $Q$, and, its contravariant dependence
is the same as $P$, while its covariant dependence is the same as $Q$.
%
The introduction rule is then a tuple $\tensorintroatXwithYandZ b s t$
with an instantiation $b$ of the quantified variable $\beta$. Note
that the use of $d^+\Psi_s$ in the rule could be replaced by
$d^-\Psi_t$ since they must be equal by the side condition $\Phi =
\jnctx \Psi_s \Psi_t$.
%
The elimination rule says to construct a transformation taking a pair
as input it is sufficient to construct one that takes in a $P$ and $Q$
for any $\beta$, directly corresponding to the adjoint relationship
between the tensor and the right hom.
%
Finally, we have the expected $\beta\eta$ equations.

\begin{figure}
  \begin{mathpar}
    \inferrule*[right=Unit-Formation]
    {\Gamma \vdash \cat C \isaCat\and
     \Gamma \pipe d^-\Xi \vdash a_1 : \cat C\and \Gamma \pipe d^+ \Xi \vdash a_2 : \cat C}
    {\Gamma \pipe \Xi \vdash \punitinXfromYtoZ {\cat C} {a_1} {a_2}}

@@

    \inferrule*[right=Unit-Introduction]
    {\Gamma\pipe\beta: \cat D \vdash a : \cat C}
    {\Gamma \pipe\cdot \vdash \punitrefl a : \punitinXfromYtoZ {\cat C} a a}

    \inferrule*[right=Unit Elim]{
      \Gamma\vdash Q : \harr {\cat C} {\cat C}\and
      \Gamma\vdash t : \pendallXdotY {\alpha:\cat C} {\harrapp Q \alpha \alpha}}
    {\Gamma \vdash \punitelimtoYwithkontZ Q N :
      \pendallXdotY {\alpha_1} {\homrallXYtoZ {\alpha_2} {\punitinXfromYtoZ {\cat C} {\alpha_1}{\alpha_2}} {\harrapp Q {\alpha_1}{\alpha_2}}}
      }

@@

    \appunary {\punitelimtoYwithkontZ Q N} a {\punitrefl a} a = \pendappXtoY N a

    \inferrule*[right=Unit$\eta$]
    {\Gamma \vdash M : \pendallXdotY {\alpha_1} \homrallXYtoZ {\alpha_2} {\punitinXfromYtoZ {\cat C} {\alpha_1} {\alpha_2}} {\harrapp Q {\alpha_1} {\alpha_2}}}
    {M = {\punitelimtoYwithkontZ Q {\appunary M \alpha {\punitrefl \alpha} \alpha}}}
    %% {}
    %% {\alpha_1 : \cat C, f : \alpha_1 \to_{\cat C} \alpha_2, \alpha_2:\cat C \vdash M : \harrapp Q {\alpha_1} {\alpha_2}}
    %% {\alpha_1 : \cat C, f : \alpha_1 \to_{\cat C} \alpha_2, \alpha_2:\cat C \vdash M =
    %%   \homrappXtoYatZ {\pendappXtoY {\punitelimtoYwithkontZ Q {\pendlambdaXdotY \alpha M[\alpha/\alpha_1,\pendappXtoY {\punitrefl C} \alpha/f,\alpha/\alpha_2]}} {\alpha_1}} f {\alpha_2}
    %% : \harrapp Q {\alpha_1}{\alpha_2}}

    \inferrule*[right=Tensor formation]
    {\Gamma \pipe d^-\Xi; \beta:D\vdash P \isaSet\and
     \Gamma \pipe \beta:D;d^+\Xi \vdash Q \isaSet}
    {\Gamma \pipe \Xi \vdash \tensorexistsXwithYandZ {\beta:D} P Q}

@@

  %% \inferrule*[right=Tensor introduction]
  %% {\Gamma \vdash P : \harr C D \and
  %%   \Gamma \vdash Q : \harr D E}
  %% {\Gamma\pipe\cdot\vdash \tensorintroofXandY P Q : \pendallXdotY {\alpha} \homrallXYtoZ \beta {\harrapp P \alpha \beta} {\homrallXYtoZ \gamma {\harrapp Q \beta \gamma} {(\tensorexistsXwithYandZ \beta {\harrapp P \alpha \beta}{\harrapp Q \beta \gamma})}}}

  \inferrule*[right=Tensor introduction]
  {\Phi = \Psi_s \jnctx \Psi_t \and
   \Gamma \pipe d^+\Psi_s \vdash b : \cat D\and
   \Gamma \pipe \Psi_s \vdash s : P[b/\beta] \and
   \Gamma \pipe \Psi_t \vdash t : Q[b/\beta]}
  {\Gamma\pipe\Phi\vdash \tensorintroatXwithYandZ b s t : \tensorexistsXwithYandZ {\beta : \cat D} P Q}

@@

  \inferrule*[right=Tensor-elim]
  {\Gamma \pipe \vdash P : \harr {\cat C} {\cat D}\and\Gamma \vdash Q : \harr {\cat D}{\cat E}\and
   \Gamma \vdash R : \harr {\cat C}{\cat E}\and
   \Gamma \vdash N : \hombinary \alpha {\harrapp P \alpha \beta} \beta {\harrapp Q \beta \gamma} \gamma {\harrapp R \alpha \gamma}}
  {\Gamma \vdash \tensorelimfromWXtoYwithkontZ P Q R N : \homunary \alpha {(\tensorexistsXwithYandZ \beta {\harrapp P \alpha \beta} {\harrapp Q \beta \gamma})} {\gamma} {\harrapp R \alpha \gamma}}

@@

  \inferrule*[right=Tensor $\beta$]
  {}
  {\appunary {\tensorelimfromWXtoYwithkontZ P Q R N} a {\tensorintroatXwithYandZ b s t} c = \appbinary N a s b t c}
  %% {\homrappXtoYatZ {\pendappXtoY{\tensorelimfromWXtoYwithkontZ P Q R N} A} {(\homrappXtoYatZ{\homrappXtoYatZ{\pendappXtoY {\tensorintroofXandY P Q} A} f B} g G)} G =
  %%   \homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY N A} f B} g G}

@@

  \inferrule*[right=Tensor$\eta$]
  {\Gamma \vdash N : \homunary \alpha {(\tensorexistsXwithYandZ \beta {\harrapp P \alpha \beta}{\harrapp Q \beta \gamma})} \gamma {\harrapp R \alpha \gamma}}
  {N = \tensorelimfromWXtoYwithkontZ P Q R {\lambdabinary \alpha x \beta y \gamma \appunary N \alpha {\tensorintroatXwithYandZ \beta x y} \gamma}}

  %% \inferrule*[right=Tensor$\eta$]
  %% {\alpha : \cat C, x : \tensorexistsXwithYandZ \beta P Q, \gamma : \cat E \vdash M : R}
  %% {\alpha : \cat C, x : \exists \beta. Q \odot R, \gamma : \cat E \vdash M =
  %%   \homrappXtoYatZ {\pendappXtoY{\tensorelimfromWXtoYwithkontZ P Q R
  %%       {\pendlambdaXdotY \alpha {\homrlambdaXatYdotZ p \beta {\homrlambdaXatYdotZ q \gamma {M[\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {\tensorintroofXandY P Q} \alpha} p \beta} q \gamma/x]}}}}
  %%     } \alpha} x \gamma : S}
  \end{mathpar}  
  \caption{Unit, Tensor of Profunctors}
\end{figure}

%% Next, we can internalize the $,\alpha,$ as a positive connective as a
%% kind of ``existentially quantified tensor product'', abstracting the
%% composition of profunctors.

%% The ordinary formula for this is that for profunctors $R : C^o \times
%% D \to \Set$ and $S : D^o \times E \to \Set$, the composite is given by
%% a coend:

%% \[ \int^{d \in D} R(-,d) \otimes S(d,=) \]

%% We will instead use slightly more logical notation, writing the coend as
%% an existential quantifier, justified by the analogy between the two
%% and the structure of the introduction/elimination forms in our syntax
%% in particular.
%% \begin{figure}
%% \begin{mathpar}
%% \end{mathpar}
%% \caption{Tensor Primitive Rules}
%% \label{fig:tensor}
%% \end{figure}

Semantically, the tensor product is the composition of profunctors,
which is typically written as a coend
\[ \int_{\beta} P(-,\beta) \times Q(\beta,=) \]

@@

%% \subsection{Equalizers}

%% To encode the concept of internal hom of profunctors, kan extensions
%% and (co-)limits, we need to define the \emph{equalizer} of horizontal
%% arrows. Specifically, the universal property that we need is an
%% equalizer in the category whose objects are horizontal arrows (for
%% fixed endpoints) and whose arrows are the 2-cells.

%% \begin{figure}
%%   \begin{mathpar}
%%     \inferrule*[right=Equalizer-formation]
%%     {\Gamma \pipe \cdot \vdash P : \harr\, C\,D\and
%%      \Gamma \pipe \cdot \vdash Q : \harr\, C \, D\\\\
%%      \Gamma \pipe \cdot \vdash \phi : \homunary \alpha {\harrapp P \alpha \beta} \beta {\harrapp Q \alpha \beta}\and
%%      \Gamma \pipe \cdot \vdash \psi : \homunary \alpha {\harrapp P \alpha \beta} \beta {\harrapp Q \alpha \beta}\and}
%%     {\Gamma \pipe \cdot \vdash \textrm{Equalizer}(C,D,P,\phi,\psi) : \harr\, C\, D}\and

%%     \inferrule
%%     {\Gamma \pipe \Phi \vdash M : \harrapp{\textrm{Equalizer}(C,D,P,\phi,\psi)} A B}
%%     {\Gamma \pipe \Phi \vdash \textrm{forget}(M) : \harrapp P A B}

%%     \inferrule
%%     {\Gamma \pipe \Phi \vdash M : \harrapp P A B \and \appunary \phi A M B = \appunary \psi A M B}
%%     {\Gamma \pipe \Phi \vdash \textrm{equalizing}_{\phi,\psi}(M) : \harrapp{\textrm{Equalizer}(C,D,P,\phi,\psi)} A B}

%%     TODO: obvious equations
%%   \end{mathpar}
%%   \caption{Equalizers (Logical Framework style)}
%% \end{figure}

%% \begin{figure}
%%   \begin{mathpar}
%%     \inferrule*[right=Equalizer-formation]
%%     {\Gamma \pipe \Xi \vdash P \isaSet\and\Gamma \pipe \Xi \vdash Q \isaSet\and
%%       \Gamma\pipe d^-\Xi, p : P, d^+ \Xi \vdash \phi : Q\and
%%       \Gamma\pipe d^-\Xi, p : P, d^+ \Xi \vdash \psi : Q }
%%     {\Gamma \pipe \Xi \vdash \equalizeVofWbyXeqYatZ p P \phi \psi Q : \harr\, C\, D}\and

%%     \inferrule*[right=Equalizer-formation]
%%     {\Gamma \pipe \alpha:C;\beta:D \vdash P \isaSet\and\Gamma \pipe \alpha:C;\beta:D \vdash Q \isaSet\and
%%       \Gamma \pipe d^-\Xi \vdash A : C\and
%%       \Gamma \pipe d^+\Xi \vdash B : D\\\\
%%       \Gamma\pipe \cdot \vdash \phi : \homunary \alpha P \beta Q\and
%%       \Gamma\pipe \cdot \vdash \psi : \homunary \alpha P \beta Q
%%     }
%%     {\Gamma \pipe \Xi \vdash \equalizeVofWbyXeqYatZ p {P[A/\alpha;B/\beta]} {\appunary \phi A p B} {\appunary \psi A p B} {Q[A/\alpha;B/\beta]} : \harr\, C\, D}\and

%%     \inferrule
%%     {\Gamma \pipe \Phi \vdash M : \equalizeVofWbyXeqYatZ p P \phi \psi Q}
%%     {\Gamma \pipe \Phi \vdash \pi_{eq}\,M : P}

%%     \inferrule
%%     {\Gamma \pipe \Phi \vdash M : \harrapp P A B \and \appunary \phi A M B = \appunary \psi A M B}
%%     {\Gamma \pipe \Phi \vdash \textrm{equalizing}_{\phi,\psi}(M) : \harrapp{\textrm{Equalizer}(C,D,P,\phi,\psi)} A B}

%%     TODO: equations
%%   \end{mathpar}
%%   \caption{Equalizers (Natural Deduction Style)}
%% \end{figure}

%% \subsection{Products}

%% We add in parallel two different notions of product: products of sorts
%% and products of spans, which have the universal property in their
%% respective categories.

%% \begin{figure}
%%   \begin{mathpar}
%%     \inferrule{}{\Gamma \vdash 1 : \Sort}\and
%%     \inferrule*{}{\Gamma\pipe\alpha:C \vdash () : 1}\and
%%     \inferrule*{\Gamma\pipe\alpha:C \vdash A : 1}{A = ()}
%%     \inferrule{}{\Gamma\pipe \Xi \vdash 1 \isaSet}\and
%%     \inferrule{}{\Gamma\pipe \Phi \vdash () :  1}\and
%%     \inferrule{\Gamma\pipe \Phi \vdash M :  1}{M = ()}\\\\

%%     \inferrule{\Gamma \vdash C : \Sort \and \Gamma \vdash D : \Sort}
%%               {\Gamma \vdash C \times D : \Sort}\and

%%     \inferrule{\forall i \in \{1,2\}.~ \Gamma \pipe \alpha:C \vdash A_i : D_i}{\Gamma \pipe \alpha:C \vdash (A_1,A_2) : D_1 \times D_2}

%%     \inferrule{\Gamma \pipe \alpha:C \vdash A : D_1 \times D_2}{\Gamma \pipe \alpha:C \vdash \pi_i A : D_i}

%%     \pi_i(A_1,A_2) = A_i\\
%%     \inferrule{\Gamma\pipe \alpha:C \vdash A :  D_1\times D_2}{A = (\pi_1 A,\pi_2 A)}\\\\

%%     \inferrule{\Gamma\pipe\Xi \vdash R \isaSet \and \Gamma\pipe\Xi \vdash S\isaSet}
%%               {\Gamma\pipe\Xi \vdash R \times S\isaSet}\and

%%     \inferrule{\forall i \in \{1,2\}.~ \Gamma \pipe\Phi \vdash M_i : R_i}{\Gamma \pipe\Phi \vdash (M_1,M_2) : R_1 \times R_2}

%%     \inferrule{\Gamma \pipe\Phi\vdash M : R_1 \times R_2}{\Gamma \pipe\Phi \vdash \pi_i M : R_i}
